---
title: "Project title"
subtitle: "Proposal"
format: html
editor: visual
---

```{r}
#| label: load-pkgs

library(tidyverse)
```

# Data 1

## Introduction and data

This data came from Equity in Athletics Data Analysis (EADA). It was collected through mandatory reports from post-secondary institutions that receive Title IX funding. The data was collected from 2020 and 2021. The data is looking as equity in college athletics by comparing different aspects of male and female sports. This specific dataset is looking at all schools in the Atlantic Coast Conference. Data was collected on multiple variables including the number of male and female athletes, the number of male and female coaches, and the revenue of male and female sports. There is also information on coaching salaries. There are no ethical concerns we foresee arising.

## Research question

Based on schools in the ACC, are the expenses for male and female sports proportional to the number of male and female athletes? Is there a difference in these proportionalities based on whether the school is public or private? It would be interesting to look at the differences between the amount of money spent by male sports vs female sports and how this compares to the number of male and female athletes. While this discrepancy has gotten better in recent years, we hypothesize that the expenses for male sports compared to female sports will be disproportionately high given the number of male and female athletes. This is important because Title IX works to provide equity in college athletics, and there has been a great deal of progess, but there is more progress that can be made. There will be quantitative variables when looking at expenses of teams and number of athletes. There will also be categorical variables when looking at the different schools (private vs public) and teams (female vs male).

## Literature

https://www.tandfonline.com/doi/pdf/10.1080/00221546.2006.11778925?casa_token=WIU1GCI895sAAAAA:w1FWFtPuxYNANBTPOiTttNFZ5up_7Cb9aiQqxKUfIkIAt_puuRYouq5c5HSirIe1idZC8qSHlot2jw

Gender Equity in Intercollegiate Athletics: Determinants of Title IX Compliance

This article analyzes the progress of Title IX in athletics from the late 1990s to the early 2000s. Their resreach looked at the proportionality gap between the number of male and female athletes. The results show that there was improvement in this gap overall. However, certain institutions had a larger gap including institutions in the south and institutions with football teams. Our questions looks to expand on this research and look at more recent numbers on equity in college athletics.

## Glimpse of data

```{r}
#| label: load-data-1

equity <- read.csv("data/ACC_2020_2021.csv")
equity |>
  slice(1:5)
```

Full glimpse was very long so just sliced the first 5 rows.

# Data 2

## Introduction and data

-   Identify the source of the data.

-   State when and how it was originally collected (by the original data curator, not necessarily how you found the data).

-   Write a brief description of the observations.

Federal law requires spills, explosions, and other events involving the spread of toxic chemicals to be reported to the Pipeline and Hazardous Materials Safety Administration (PHMSA). Whether reported by individuals or companies, PHMSA shows these reports on their online portal. However, the data collection of the agency is not idea for analysis. The Data Liberation Project fetches the data from the portal and organizes the data in a way that can be used for statistical analysis. Granted, the amount of data collected is overwhelming, so there will be a focus on data specifically from February 2023. Data seems to show locations and descriptions of the incidents, along with a plethora of other information.

Research question

-   A well formulated research question. (You may include more than one research question if you want to receive feedback on different ideas for your project. However, one per data set is required.)

What chemicals are most often found in hazardous spills? How frequently do spills occur? In what situation are these spills most likely to take place.

-   A description of the research topic along with a concise statement of your hypotheses on this topic.

The primary goal of this research is to discover causes of hazardous spills and who is usually responsible for them.

-   Identify the types of variables in your research question. Categorical? Quantitative?

The majority of these variables are categorical because spills depend on a myriad of factors. However, percentages may come into play when considering chemical analysis

## Literature

-   Find one published credible article on the topic you are interested in researching.

    https://abcnews.go.com/US/feds-norfolk-southerns-handling-additional-reported-hazmat-concern/story?id=97723774

-   Provide a one paragraph summary about the article.

The article highlights one of the most high profile chemical spills of the last few years. A train derailment in East Palestine, Ohio caused an extreme amount of toxic chemicals to be released into the the surrounding area. Beyond that, there seems to be concerns about additional hazards that were previously unreported according to the federal government. The event has become national news and has brought into question regulations surrounding toxic chemical transport and potentially neglect on behalf of the company Norfolk Southern

-   In 1-2 sentences, explain how your research question builds on / is different than the article you have cited.

The research question takes a larger look at hazardous spill incidents and how they are potentially caused. While the East Palestine train derailment is the most infamous example as of late, these spills are actually much more common than many might realize.

## Glimpse of data

```{r}
#| label: load-data-2

feb_chem_spill_data <- read.csv("data/fetched/2023-02.csv") 
```

-   Running into a slight issue here with opening the file

# Data 3

## Introduction and data

-   Identify the source of the data.

-   State when and how it was originally collected (by the original data curator, not necessarily how you found the data).

-   Write a brief description of the observations.

## Research question

-   A well formulated research question. (You may include more than one research question if you want to receive feedback on different ideas for your project. However, one per data set is required.)

-   A description of the research topic along with a concise statement of your hypotheses on this topic.

-   Identify the types of variables in your research question. Categorical? Quantitative?

## Literature

-   Find one published credible article on the topic you are interested in researching.

-   Provide a one paragraph summary about the article.

-   In 1-2 sentences, explain how your research question builds on / is different than the article you have cited.

## Glimpse of data

```{r}
#| label: load-data-3

# add code here
```
